### 2019-01-26 assignments/2/assignment_2.Rmd
Initial commit

---
title: 'EEOB563 -- Assignment #2'
author: "Dennis Psaroudakis"
date: "January 26th 2019"
output:
  pdf_document:
    dev: pdf
    latex_engine: lualatex
    md_extensions: +link_attributes+multiline_tables
    pandoc_args: --listings
mainfontoptions: Numbers=OldStyle
monofont: Linux Libertine Mono O
mainfont: Linux Libertine O
papersize: letter
sansfont: Linux Biolinum O
language: en-US
...

## 2

### 2019-01-27 assignments/2/assignment_2.Rmd
Assignment 2: part 2 done
Finish HW2

@@
 
-## 2
+## 3
+GitHub repo is https://github.com/Thyra/EEOB563
+
+## 4
+a) The dinosaur sequence is highly similar to the cloning vector pAgaL as well as some other cloning vectors.
+b) The highest rated sequence is erythroid transcription factor from *Gallus gallus*. It is only 96% identical, there are some differences:
+```
+Query  121   MEFVALGGPDAGSPTPFPDeagaflglgggerteaggllaSYPPSGRVSLVPWADTGTLG  300
+             MEFVALGGPDAGSPTPFPDEAGAFLGLGGGERTEAGGLLASYPPSGRVSLVPWADTGTLG
+Sbjct  1     MEFVALGGPDAGSPTPFPDEAGAFLGLGGGERTEAGGLLASYPPSGRVSLVPWADTGTLG  60
+
+Query  301   TPQWVPPATQMEPPHYLEllqpprgspphpssgpllplssgpppCEARECVMARKNCGAT  480
+             TPQWVPPATQMEPPHYLELLQPPRGSPPHPSSGPLLPLSSGPPPCEARECV    NCGAT
+Sbjct  61    TPQWVPPATQMEPPHYLELLQPPRGSPPHPSSGPLLPLSSGPPPCEARECV----NCGAT  116
+
+Query  481   ATPLWRRDGTGHYLCNWASACGLYHRLNGQNRPLIRPKKRLLVSKRAGTVCSHERENCQT  660
+             ATPLWRRDGTGHYLCN   ACGLYHRLNGQNRPLIRPKKRLLVSKRAGTVCS    NCQT
+Sbjct  117   ATPLWRRDGTGHYLCN---ACGLYHRLNGQNRPLIRPKKRLLVSKRAGTVCS----NCQT  169
+
+Query  661   STTTLWRRSPMGDPVCNNIHACGLYYKLHQVNRPLTMRKDGIQTRNRKVsskgkkrrppg  840
+             STTTLWRRSPMGDPVCN   ACGLYYKLHQVNRPLTMRKDGIQTRNRKVSSKGKKRRPPG
+Sbjct  170   STTTLWRRSPMGDPVCN---ACGLYYKLHQVNRPLTMRKDGIQTRNRKVSSKGKKRRPPG  226
+
+Query  841   ggnpsatagggapmggggdpsmpppppppaaappQSDALYALGPVVLSGHFLPfgnsggf  1020
+             GGNPSATAGGGAPMGGGGDPSMPPPPPPPAAAPPQSDALYALGPVVLSGHFLPFGNSGGF
+Sbjct  227   GGNPSATAGGGAPMGGGGDPSMPPPPPPPAAAPPQSDALYALGPVVLSGHFLPFGNSGGF  286
+
+Query  1021  fgggaggYTAPPGLSPQI  1074
+             FGGGAGGYTAPPGLSPQI
+Sbjct  287   FGGGAGGYTAPPGLSPQI  304
+```
+
+## 5
+These sequences might be homologous (I don't think I can be 100% sure just from the blast result) and if they are, they are more specificially orthologous because they are in different species as opposed to in different genome locations within the same species.
+
+## 6
+I let mafft choose the strategy (auto) and it chose L-INS-i, probably because there were so few sequences and not too long.
+
+![Tree created by mafft](./mafftTree.png){width=65%}}
+
+This tree is a guide tree so it is only a rough clustering of the species, not the final phylogenic tree. The actual algorithm to create the phylogenetic tree is run later.

### 2019-01-27 assignments/2/assignment_2.Rmd
Rendered Assignment 2, only pen and paper part missing

@@
 language: en-US
+titlepage: true
 ...
@@
 b) The highest rated sequence is erythroid transcription factor from *Gallus gallus*. It is only 96% identical, there are some differences:
+
 ```
@@
 
-![Tree created by mafft](./mafftTree.png){width=65%}}
+![Tree created by mafft](./mafftTree.pdf){width=95%}
 

### 2019-02-21 assignments/3/assignment_3.Rmd
Assignment 3, part I

---
title: 'EEOB563 -- Assignment #3'
author: "Dennis Psaroudakis"
date: "February 20th 2019"
output:
  pdf_document:
    dev: pdf
    latex_engine: lualatex
    md_extensions: +link_attributes+multiline_tables
    pandoc_args: --listings
mainfontoptions: Numbers=OldStyle
monofont: Linux Libertine Mono O
mainfont: Linux Libertine O
papersize: letter
sansfont: Linux Biolinum O
language: en-US
titlepage: true
...

## 3
c) I did an alltree search on the sequences and this is the result

```
Exhaustive search settings:
  Optimality criterion = parsimony
    Character-status summary:
      Of 30 total characters:
        All characters are of type 'unord'
        All characters have equal weight
        22 characters are constant (proportion = 0.733333)
        6 variable characters are parsimony-uninformative
        Number of parsimony-informative characters = 2
    Gaps are treated as "missing"
  Initial 'Maxtrees' setting = 100
  Branches collapsed (creating polytomies) if maximum branch length is zero
  'MulTrees' option in effect
  No topological constraints in effect
  Trees are unrooted

Exhaustive search completed:
  Number of trees evaluated = 2027025
  Score of best tree found = 11
  Score of worst tree found = 16
  Number of trees retained = 3
  Time used = 0.54 sec (CPU time = 0.54 sec)
```
So the best score is 11, according to PAUP. I don't know why it is so different from my score, but I guess it just uses a different algorithm and/or different values in the algorithm. The best tree it found looks like this:

```
Tree 3 (rooted using default outgroup)

Tree length = 11
Consistency index (CI) = 1.0000
Homoplasy index (HI) = 0.0000
CI excluding uninformative characters = 1.0000
HI excluding uninformative characters = 0.0000
Retention index (RI) = 1.0000
Rescaled consistency index (RC) = 1.0000

/----------------------------------------------------------------------------------------------------------------------------------------- Dolphin
|
|                                                          /--------------------------------------- Hippopotamus
|                                                          |
|                                                          |                                      /--------------------------------------- Cow
|                   /-------------------------------------13                                      |
|                   |                                      +-------------------------------------11--------------------------------------- Giraffe
|                   |                                      |                                      |
|                   |                                     12                                      \--------------------------------------- Goat
|                   |                                      |
|                   |                                      \--------------------------------------- Sheep
|                   |
\------------------14 Camel
                    |
                    + Sperm whale
                    |
                    + Blue whale
                    |
                    \ Pig
```
Please look at my paup log file for additional info, here is also the FASTA file I wrote and the Nexus file I generated from it.

### 2019-02-21 assignments/3/assignment_3.Rmd
Started Assignment 3 Part II

@@
 ```
-Please look at my paup log file for additional info, here is also the FASTA file I wrote and the Nexus file I generated from it.
+Please look at my [paup log](https://github.com/Thyra/EEOB563/blob/master/assignments/3/3c.log) file for additional info, here is also the [FASTA file](https://github.com/Thyra/EEOB563/blob/master/assignments/3/3.fa) I wrote and the [Nexus file](https://github.com/Thyra/EEOB563/blob/master/assignments/3/3.nxs) I generated from it.
+
+# Part II
+I aligned the sequences with Pal2Nal (vertebrate table, output format FASTA, I told it to remove gaps, otherwise elephant and human were shorter than the others.
+Otherwise I left the standard options), I got a lot of warnings that I does not correspond to ATA but it seems like it worked ([result file](https://github.com/Thyra/EEOB563/blob/master/assignments/3/cob_aligned.fa)).
+I converted it into a [Nexus file](https://github.com/Thyra/EEOB563/blob/master/assignments/3/cob_aligned.nxs).
+
+## 4
+Please look at the complete log file of the analysis [here](https://github.com/Thyra/EEOB563/blob/master/assignments/3/4.log).
+All the trees that were found had a very similar score and only marginally different topologie, the run with 100 reps was the most different from the rest.
+It reminds me of the trees I did by hand where I also found the best tree very early.
+So either this is just the global optimum and very easy to find or it is a local optimum and none of the methods were thorough enough to find the global optimum.
+
+## 5

### 2019-02-21 assignments/3/assignment_3.Rmd
Rendered assignment

@@
     md_extensions: +link_attributes+multiline_tables
-    pandoc_args: --listings
+    pandoc_args: --listings --colorlinks
 mainfontoptions: Numbers=OldStyle
@@
 ## 5
+I used the mafft alignment for this one because I somehow didn't trust the Pal2Nal result. No, the trees are not the same:
+
+![Majority consensus tree](majTree.png){width=60%}
+
+![Strict consensus tree](strictTree.png){width=60%}
+
+Their topology is different and even more importantly the majority consensus tree contains distances while the strict consensus tree does not.

### 2019-02-28 assignments/4/assignment_4.Rmd
Rendered assignment 4

---
title: 'EEOB563 -- Assignment #4'
author: "Dennis Psaroudakis"
date: "February 27th 2019"
output:
  pdf_document:
    dev: pdf
    latex_engine: lualatex
    md_extensions: +link_attributes+multiline_tables
    pandoc_args: --listings --colorlinks
mainfontoptions: Numbers=OldStyle
monofont: Linux Libertine Mono O
mainfont: Linux Libertine O
papersize: letter
sansfont: Linux Biolinum O
language: en-US
titlepage: true
...

## 4.
The generated matrices and trees can be found [here](https://github.com/Thyra/EEOB563/tree/master/assignments/4/4). I chose Platypus (20) as the outgroup, this is the generated tree:

```
(((((Elephant,Sloth,(Flying_fox,(Whale,Pig),(Human,Cat),Armadillo,
Hedgehog,(Rabbit,Aardvark),Dog,Tenrec,Tree_shrew),Mouse),(Opossum,
Bandicoot,Wallaby),Echidna),Platypus);))))
```

![Strict consensus tree](4/consensus_tree.png){width=60%}

## 5.
I used the Jukes-Cantor distance model, just because I used it for Part I so I knew most about it. This is the generated consensus tree from all the bootstrapped alignments:

```
((((((((((Cat:1.00,Rabbit:1.00):1.00,Mouse:1.00):1.00,Dog:1.00):1.00,((Elephant:1.00,
Sloth:1.00):1.00,Human:1.00):1.00):1.00,(((Armadillo:1.00,Whale:1.00):1.00,Pig:1.00):1.00,
(Flying_fox:1.00,Tree_shrew:1.00):1.00):1.00):1.00,Aardvark:1.00):1.00,(Hedgehog:1.00,
Tenrec:1.00):1.00):1.00,(Opossum:1.00,(Wallaby:1.00,Bandicoot:1.00):1.00):1.00):1.00,
Platypus:1.00):1.00,Echidna:1.00);
```

![Majority consensus tree](5/consensus_tree.png){width=60%}

## 8.
Again, I used the Jukes-Cantor matrix. In my case, the two trees basically were the same, only the distances were slightly different:

Without NNI/SPR:
```
(((((Cat:0.114165981250,Human:0.162999018750):0.010559694336,((((Hedgehog:0.178496817708,Tenrec:0.157865182292):0.012008114955,(Mouse:0.121806644097,(((Bandicoot:0.115205470588,Opossum:0.129977529412):0.001976093750,Wallaby:0.119019406250):0.017233485577,(Echidna:0.102001583333,Platypus:0.112883416667):0.066429764423):0.010216355903):0.004562978795):0.005054285156,(Armadillo:0.119819296875,(Sloth:0.152378928571,Elephant:0.163353071429):0.012417703125):0.006888199219):0.002261122559,(Rabbit:0.131889476562,Aardvark:0.120376523438):0.005459932129):0.002273915039):0.003076952637,(Dog:0.127646692383,Tree_shrew:0.148635307617):0.001928684082):0.011074565918,Flying_fox:0.117479167614):0.013291332386,Pig:0.093116691667,Whale:0.119435308333);
```

With NNI/SPR:
```
(((((Cat:0.112028724609,Human:0.165136275391):0.009859859863,((((Hedgehog:0.177491945312,Tenrec:0.158870054687):0.009443020020,(Mouse:0.123529160645,(((Bandicoot:0.113613665527,Opossum:0.131569334473):0.004081156738,Wallaby:0.116914343262):0.018083876465,(Echidna:0.105270403320,Platypus:0.109614596680):0.065579373535):0.008493839355):0.007128073730):0.005542280762,(Armadillo:0.119178175781,(Sloth:0.143488730469,Elephant:0.172243269531):0.013058824219):0.006400203613):0.002117241699,(Rabbit:0.130965949219,Aardvark:0.121300050781):0.005603812988):0.002973749512):0.003076952637,(Dog:0.127646692383,Tree_shrew:0.148635307617):0.001928684082):0.011074565918,Flying_fox:0.117013964355):0.013756535645,Pig:0.087929280762,Whale:0.124622719238);
```
Left: Without NNI/SPR, right: with NNI/SPR

![Without NNI/SPR](8/without.png "Without NNI/SPR"){width=45%} ![With NNI/SPR](8/with.png "With NNI/SPR"){width=45%}

I may have done something wrong, you can check my parameters and input [here](https://github.com/Thyra/EEOB563/tree/master/assignments/4/8).

### 2019-03-12 assignments/5/Assignment_5.Rmd
Assignment 5.3 done

3
F84 is not available in raxml, so I'm using F81
Raxml: Final LogLikelihood -54568.308901
FastMe: Final LogLikelihood -54674.432495

4

### 2019-03-13 assignments/5/Assignment_5.Rmd
Did Assignment 5.4

@@
-3
-F84 is not available in raxml, so I'm using F81
-Raxml: Final LogLikelihood -54568.308901
-FastMe: Final LogLikelihood -54674.432495
+## 3
+F84 is not available in raxml, so I'm using F81.
+The trees are not the same (see following tanglegram), they also have different log-likelihood scores:
 
-4
+```
+Raxml:  -54568.308901
+FastMe: -54674.432495
+```
+![Tanglegram made with Dendroscope](3/Tanglegram.png)
+
+## 4
+These are the two commands I used:
+
+```
+raxml-ng --search --msa ../alignment.phy --model model_codons --prefix codons --seed 12 --brlen scaled
+raxml-ng --search --msa ../alignment.phy --model model_genes --prefix genes --seed 12 --brlen scaled
+```
+
+The two models looked like this:
+
+````
+model_codons:
+  GTR+G+FO, COBX=1-4482/3
+
+model_genes:
+  GTR+G+FO, cob=1-1248
+  GTR+G+FO, cox1=1249-2808
+  GTR+G+FO, cox2=2809-3543
+  GTR+G+FO, cox3=3544-4482
+```
+
+These are the log-likelihoods for the best tree:
+
+```
+Codons: -45373.634679
+Genes:  -45270.430490
+```
+
+They are much better than the original tree BUT the evolution models I used are different so I'm not sure if it's directly comparable.
+
+Anyway, here is the difference between the Gene partitioned tree and the original RaXML tree, they are very similar except for the rabbit:
+
+![Tanglegram made with Dendroscope](4/Genes_vs_3_raxml.png)
+
+## 5
+Unfortunately I can't do this part right now because HPC is having problems with slurm or something (I copied you on an email about this).
+I will resubmit this assignment including this part when it's fixed.

### 2019-03-13 assignments/5/Assignment_5.Rmd
Rendered Assignment 5 without last part

@@
+---
+title: 'EEOB563 -- Assignment #5'
+author: "Dennis Psaroudakis"
+date: "March 12th 2019"
+output:
+  pdf_document:
+    dev: pdf
+    latex_engine: lualatex
+    md_extensions: +link_attributes+multiline_tables
+    pandoc_args: --listings
+    template: ../template.tex
+mainfontoptions: Numbers=OldStyle
+monofont: Linux Libertine Mono O
+mainfont: Linux Libertine O
+papersize: letter
+sansfont: Linux Biolinum O
+language: en-US
+titlepage: true
+...
+
 ## 3
@@
 ```
-![Tanglegram made with Dendroscope](3/Tanglegram.png)
+![Tanglegram made with Dendroscope (zoom in)](3/Tanglegram.png)
+
+\clearpage
 
@@
 
-````
+```
 model_codons:
@@
 
-![Tanglegram made with Dendroscope](4/Genes_vs_3_raxml.png)
+![Tanglegram made with Dendroscope (zoom in)](4/Genes_vs_3_raxml.png)
+
+\clearpage
 
 ## 5
-Unfortunately I can't do this part right now because HPC is having problems with slurm or something (I copied you on an email about this).
-I will resubmit this assignment including this part when it's fixed.
+I had problems with slurm so I couldn't start it until Tuesday evening, so it's still running.
+I will resubmit this assignment tomorrow with part 5 when it is done.
+I am expecting a runtime of $\ge$ 10 hours from what the others have told me, I don't remember the runtimes from the lab anymore.
+This is my slurm script:
+
+```bash
+#!/bin/bash
+
+#SBATCH --time=10:30:00   # walltime limit (HH:MM:SS)
+#SBATCH --nodes=1   # number of nodes
+#SBATCH --ntasks-per-node=16   # 16 processor core(s) per node 
+#SBATCH --job-name="5.5_assignment"
+#SBATCH --mail-user=dpsaroud@iastate.edu   # email address
+#SBATCH --mail-type=BEGIN
+#SBATCH --mail-type=END
+#SBATCH --mail-type=FAIL
+
+# LOAD MODULES, INSERT CODE, AND RUN YOUR PROGRAMS HERE#!/bin/bash
+cd /home/dpsaroud/EEOB563/assignments/5/5
+
+/home/dpsaroud/bin/raxml-ng --all --msa ../alignment.phy --model GTR+G --seed 12 --threads 16 --bs-metric fbp --bs-trees 1000
+```

### 2019-03-13 assignments/5/Assignment_5.Rmd
Finished 5.5

@@
 ## 5
-I had problems with slurm so I couldn't start it until Tuesday evening, so it's still running.
-I will resubmit this assignment tomorrow with part 5 when it is done.
-I am expecting a runtime of $\ge$ 10 hours from what the others have told me, I don't remember the runtimes from the lab anymore.
+I had expected a runtime of $\ge$ 10 hours from what the others have told me, I don't remember the runtimes from the lab anymore. In reality, it only took about 6:30 hours though, so I was lucky.
 This is my slurm script:
@@
 ```
+
+Here is the tree, finally the tree is further away from the animals:
+
+![ML tree with bootstrap support values](5/BootstrapSupport.png)

### 2019-03-27 final_project/Outline.Rmd
Add final project outline

---
title: "Gene Function Phylogeny"
author: "Dennis Psaroudakis"
date: "March 26th 2019"
keep_md: yes
output:
  tufte::tufte_handout:
    latex_engine: xelatex
    citation_package: natbib
  tufte::tufte_html: default
bibliography: outline-resources/bib.bib
link-citations: yes
classoption: justified
language: en-US
header-includes:
  - \usepackage[osf]{libertine}
  - \citestyle{plain}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

`r newthought("In my research,")` my colleagues and I have generated (or are currently generating) genome-wide functional annotation datasets for wheat, rice, cotton, soy, barley, and several maize genome assemblies using the Gene Ontology^[http://geneontology.org/].
It would be interesting to find out whether a phylogenetic tree built on the functional annotation of the genes instead of the DNA sequence would yield a tree that is similar to the currently used tree for the evolutionary relationship between these species, i.e. whether their evolution can be retraced on a functional level.

For now this is largely explorative and it is possible that I will discover interesting but different patterns than what I expected. Maybe it's also all just nonsense, we will find out.

## Data
The functional annotation data is being generated with the GOMAP pipeline^[see https://dill-picl.org/projects/gomap for more info and https://github.com/Dill-PICL/GOMAP-singularity for the source code] which uses several different approaches to predict the functions of all genes in a given genome (publication in the works, see [@maizeGAMER] for the precursor developed on maize).
All datasets are or will be published at https://dill-picl.org/projects/gomap/gomap-datasets/ .

## Method
I am planning to build a distance matrix between all annotation sets using the hF~1~ metric defined in [@AIGO].
hF~1~ is the harmonic mean of the hierarchical precision hPr and hierarchical recall hRc^[ The hF~1~ is basically the tree version of the regular [F~1~ score](https://en.wikipedia.org/wiki/F1_score).]: 
$$\textrm{hF}_1 = 2 \cdot \frac{\textrm{hPr} \cdot \textrm{hRc}}{\textrm{hPr} + \textrm{hRc}}$$
The hierarchical precision hPr is defined as follows:
$$\textrm{hPr} = \frac{\textrm{predicted GO term} \cup \textrm{true GO term}}{\textrm{predicted GO term}}$$
and the hierarchical recall correspondingly:
\begin{marginfigure}
\includegraphics{outline-resources/hF-graphical.png}
\caption{\textbf{(Zoom in)}. Graphic of how hPr and hRc use the Gene Ontology tree to calculate the overlap of two GO terms (the ones marked with a leaf)}
\label{fig:marburg_virus}
\end{marginfigure}
$$\textrm{hRc} = \frac{\textrm{predicted GO term} \cup \textrm{true GO term}}{\textrm{true GO term}}$$
To go from single annotation hF~1~ scores, those are commonly averaged to get gene or annotation set wide hF~1~ metrics.
$$\textrm{overall hF}_1 = \sum_{i=1}^{n} \frac{\textrm{hF}_i}{n}$$

The regular use case is that you have one set of GO predictions and one set of experimentally deduced GO annotations for the same genome which you can use as a gold standard to evaluate the quality of your predictions (as we did in [@maizeGAMER]).
In my case, I only want to compare two datasets without regards to any quality.
Since hPr and hRc are both used in commutative contexts for the hF~1~ metric, I can treat either of the datasets as the gold standard and the other one as the prediction.
So my hF score between sets $A$ and $B$^[$A,B \in$ \{wheat, rice, cotton, soy, barley, maize~1~, maize~2~...\}] would be:
$$\textrm{hF}_{A,B} = 2 \cdot \frac{\frac{A \cup B}{A} \cdot \frac{A \cup B}{B}}{\frac{A \cup B}{A} + \frac{A \cup B}{B}}$$

$\textrm{hF}_{A,B}$ will range between $0$ and $1$, with $1$ being the score for two completely identical datasets.
To turn this into a distance $d_{A,B}$ I can simply subract it from $1$, making the distance between identical datasets $0$ and completely disjunct datasets $1$:
$$d_{A,B} = 1 - \textrm{hF}_{A,B}$$

Then finally I can apply Neighbor-Joining to the distance matrix I generated this way.

## Approach
In the initial attempt, I want to ignore any genes and simply look at all functions present in any gene within the genome.
That means I will combine all gene GO trees of a species into one big species-wide function tree and compare these function trees.
Possibly I will limit it to the *Molecular Function* aspect of the Gene Ontology or at least exclude *Cellular Component* because I don't expect that one plant species will have, e.g. a nucleus and the other ones don't.

Then I'll see if what I have generated makes any sense and go from there.

### 2019-04-15 final_project/create_slim_ontology.py
Built tree for final project

@@ -1,0 +1,41 @@
+# This script reads an OBO file using the obo_parser and outputs a JSON file that is then used by the crystal scripts.
+# The JSON file looks like this:
+# {
+#   "parents": {
+#     "GO:122335": ["GO:4433", "GO:5432"],
+#     "GO:122341": ["GO:1243", "GO:1532"],
+#     ...
+#   },
+#   "main_id" : {"GO:1443":"GO:532", "GO:2533":"GO:5234"}
+#   "obsolete": ["GO:1432", "GO:4322", ...]
+# }
+
+import obo_parser
+import json
+
+parents = {}
+main_id = {}
+obsolete = []
+with open("go.obo") as obofile:
+  parser = obo_parser.Parser(obofile)
+  for stanza in parser:
+    if not stanza.name == "Term":
+      continue
+
+    go_id = stanza.tags["id"][0]
+
+    if 'is_a' in stanza.tags:
+      parents[go_id] = stanza.tags['is_a']
+    else:
+      parents[go_id] = []
+
+    if 'alt_id' in stanza.tags:
+      for alt_id in stanza.tags['alt_id']:
+        main_id[alt_id] = go_id
+
+    if 'is_obsolete' in stanza.tags and stanza.tags['is_obsolete'][0] == "true":
+      obsolete.append(go_id)
+
+with open('slimGO.json', 'w') as f:
+  f.write(json.dumps({"parents":parents,"main_id":main_id,"obsolete":obsolete},indent=2))
+

### 2019-04-16 final_project/paper/Paper.Rmd
Started working on Draft

---
title: "Gene Function Phylogenetics"
author: "Dennis Psaroudakis"
date: "April 18th 2019"
keep_md: yes
output:
  tufte::tufte_handout:
    latex_engine: xelatex
    citation_package: natbib
  tufte::tufte_html: default
bibliography: resources/bib.bib
link-citations: yes
classoption: justified
language: en-US
header-includes:
  - \usepackage[osf]{libertine}
  - \citestyle{plain}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

`r newthought("Introduction, Data...")`

## Method
Starting point of our method is a functional annotation set for each genome $S_a$, consisting of tuples $(G_i, T_i)$ with $G_i \in$ Genes in genome $S_a$ and $T_i \in$ Terms in the Gene Ontology. So the annotation set maps each gene of the genome to $n$ GO terms.
\marginnote[-4em]{
\texttt{Gene\ \ \ \ \ \ \ \ \ GO Term\\
Os01g0601625	GO:0050896\\
Os01g0601625	GO:0016021\\
Os01g0601625	GO:0016301\\
Os01g0601651	GO:0003677\\
Os01g0601651	GO:0009699\\
Os01g0601651	GO:0050790\\
Os01g0601651	GO:0050794\\
Os01g0601651	GO:0050896\\
Os01g0601675	GO:0007275\\
Os01g0601675	GO:0016310\\
Os01g0601675	GO:0050789\\
...
}
}

Now we use the structure of the Gene Ontology to obtain all ancestors $A_i$ of GO term $T_i$ in the GO hierarchy and combine all these annotations and their ancestors, irrespective of the Gene they are associated with: $\bigcup_{i=1}^{x}(T_i \cup A_{i})$ The resulting set is a set of all the GO terms present in the genome annotation and their ancestors.

After this is done for all annotation sets, we introduce the Jaccard distance as a metric of (dis-)similarity between the annotation sets:

$$ \textrm{Jaccard Distance}(S_a, S_b) = 1 - \frac{|S_a \cap S_b|}{|S_a \cup S_b|} $$

\begin{marginfigure}
\includegraphics{resources/hF-graphical.png}
\caption{\textbf{(Zoom in)}. Graphic of how we use the Gene Ontology tree to calculate the overlap (= intersection) of two GO terms (the ones marked with a leaf)}
\label{fig:marburg_virus}
\end{marginfigure}

Then finally I can apply Neighbor-Joining to the distance matrix I generated this way.

### 2019-04-16 final_project/paper/Paper.Rmd
Rephrased methods part

@@
 ---
-title: "Gene Function Phylogenetics"
+title: "Genome Function Phylogenetics"
 author: "Dennis Psaroudakis"
@@
 
-`r newthought("Introduction, Data...")`
+`r newthought("Introduction, Gene Ontology, Data...")`
 
 ## Method
-Starting point of our method is a functional annotation set for each genome $S_a$, consisting of tuples $(G_i, T_i)$ with $G_i \in$ Genes in genome $S_a$ and $T_i \in$ Terms in the Gene Ontology.
-So the annotation set maps each gene of the genome to $n$ GO terms.
-\marginnote[-4em]{
+Starting point of our method are the functional annotation sets, one for each genome, which annotate every gene in the genome with one or more GO terms.
+In more mathematical terms the genome annotation set is a list of tuples $(G, T)$ with $G \in$ Genes in that genome and $T \in$ Terms in the Gene Ontology.
+\marginnote[-5em]{
 \texttt{Gene\ \ \ \ \ \ \ \ \ GO Term\\
@@
 
-Now we use the structure of the Gene Ontology to obtain all ancestors $A_i$ of GO term $T_i$ in the GO hierarchy and combine all these annotations and their ancestors, irrespective of the Gene they are associated with: $\bigcup_{i=1}^{x}(T_i \cup A_{i})$ The resulting set is a set of all the GO terms present in the genome annotation and their ancestors.
+We can use the hierarchical structure of the Gene Ontology to obtain the ancestors $A_i$ of any term $T_i$; in other words the gene $G_i$ is not just annotated with the term $T_i$ itself but also with all GO terms that are a more general statement of that term (e.g.
+any gene that is part of a metabolic process is thereby also part of a biological process).
+We do that for all terms $T$ in the dataset and combine all of the terms and their ancestors into one big genome-wide set $S$, irrespectively of the gene they were originally associated with:  $S = \bigcup_{i=1}^{x}(T_i \cup A_{i})$.
 
-After this is done for all annotation sets, we introduce the Jaccard distance as a metric of (dis-)similarity between the annotation sets:
+When this superset of annotations is created for each of the datasets, we can use the Jaccard Distance as a measure of how (dis-)similar any two sets are from each other, or in biological terms how different the two genomes are on a functional level:
 
@@
 \begin{marginfigure}
+The general idea of using the Jaccard Distance in this context is to measure the overlap of two subtrees in the GO hierarchy.
+Say, for simplicity, that we're looking at two genomes (here called Gold Standard and Prediction) that each only contain one single GO term (marked by a leaf).
+First, we add all ancestors of that leaf term to each subtree.
+Then, we determine the overlap (which corresponds to $S_a \cap S_b$), and divide the number of nodes in this overlap by the number of nodes in either of the two subtrees ($S_a \cup S_b$).
+\vspace{0.2em}
+
 \includegraphics{resources/hF-graphical.png}
-\caption{\textbf{(Zoom in)}. Graphic of how we use the Gene Ontology tree to calculate the overlap (= intersection) of two GO terms (the ones marked with a leaf)}
-\label{fig:marburg_virus}
+
+\vspace{0.5em}
+\noindent In the case of this example, the Jaccard Distance of Gold Standard and Prediction would be $1 - \frac{4}{9} = \frac{5}{9}$
 \end{marginfigure}
 
-Then finally I can apply Neighbor-Joining to the distance matrix I generated this way.
+Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm.

### 2019-04-17 final_project/paper/Paper.Rmd
Contd working on paper

@@
 
-`r newthought("Introduction, Gene Ontology, Data...")`
+`r newthought("When we build phylogenetic trees,")` we often use the sequences of certain genes as the basis for our tree.
+They are the direct substrate of evolution so that makes sense, but when we think of evolution on a macro level, we don't really think of the sequences, we think of phenotype.
+Using phenotypes for evolutionary trees is a hard thing to do though because the choice of characteristics can be kind of arbitrary (http://wiki.c2.com/?BorgesClassificationOfAnimals Chinesische Tiere (Jorge Luis Borges 1966)), because it is very far away from the material evolution actually happens on, i.e.
+DNA.
+In this paper, I tried to go a middle ground: Instead of looking at the genetic sequences or the macromolecular phenotype, I'm looking at the presence of certain biological processes, molecular functions, or cellular components, in other words: along the path of evolution, any meaningful change in an organism goes along with the development, change, or loss of such a function; what good is a change in the DNA sequence if it does not lead to a modified function of that gene?
+
+## Gene Ontology
+Historically, the function/role that a gene plays in an organism has always been described in natural language, however the researcher characterizing that gene deemed best.
+While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind these words.
+Additionally, different people will describe the same thing with different words, which has the potential for misunderstanding.
+lalala Gene Ontology
+\marginnote{"The mission of the GO Consortium is to develop an up-to-date, comprehensive, computational model of biological systems, from the molecular level to larger pathways, cellular and organism-level systems."
+
+\raggedleft --- GO Consortium (geneontology.org)}
+The function/role that a gene plays in an organism can be quite a vague expression and if you look at papers doing research on these things, you will see that every researcher describes it differently.
+
+## Data
+The data that serve as the input for this project are functional annotation datasets in the GAF file format^[\scriptsize http://geneontology.org/docs/go-annotation-file-gaf-format-2.0/] which we generated with our GOMAP pipeline^[\scriptsize https://dill-picl.org/projects/gomap].
+They are (or will be shortly) available from \scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.
 

### 2019-04-18 final_project/paper/Paper.Rmd
Continue final paper

@@
 
-## Gene Ontology
+## The Gene Ontology
 Historically, the function/role that a gene plays in an organism has always been described in natural language, however the researcher characterizing that gene deemed best.
-While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind these words.
+While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind the words.
 Additionally, different people will describe the same thing with different words, which has the potential for misunderstanding.
-lalala Gene Ontology
+
+Ontologies try to alleviate these problems by providing a strictly organized and controlled vocabulary and defined relationships between the terms, so that the same statement always means the same thing, no matter the context or the  author.
+Additionally, ontologies can be understood by computers if all relationships and terms are clearly defined.
+
+The Gene Ontology is such an ontology that describes genes by the properties of their product.
 \marginnote{"The mission of the GO Consortium is to develop an up-to-date, comprehensive, computational model of biological systems, from the molecular level to larger pathways, cellular and organism-level systems."
@@
 \raggedleft --- GO Consortium (geneontology.org)}
-The function/role that a gene plays in an organism can be quite a vague expression and if you look at papers doing research on these things, you will see that every researcher describes it differently.
+In our case these gene products are proteins, and they can be characterized in three different aspects:
+
+- What biological processes is this protein part of? (e.g. photosynthesis or autophagy)
+- What molecular functions does the protein carry out? (e.g. ethylene binding or RNA ligation)
+- What cellular component is the protein active at? (e.g. outer membrane or nucleus)
+
+Within each of these aspects, the Gene Ontology defines a huge number of terms (2,675,070 in total), that range from very general to very specific:
+
+![subtree of the *Biological Process* ontology. 
+The terms are organized in such a way that more general terms are always true for any of their more specific child terms.
+For example, any protein that is part of *Cytokinesis after Meiosis I*, is also obviously part of Cytokinesis, the Cell Cycle etc.
+That way, a gene that has been annotated with the term *Cytokinesis after Meisosis I* (leaf term), has implicitly been annotated with all of that term's parent terms as well, all the way up to the root term.](resources/go_hierarchy_example.png){width=60%}
+
+\noindent When proteins are annotated with these terms instead of just natural language, we can now computationally answer some interesting questions, such as:
+
+- How similar in function is protein A to protein B? (One answer would be: How many steps in the GO graph do I need from term A to B? The fewer steps, the more similar the function)
+- If protein A is involved in Biological Process XYZ, what other proteins are involved in that same process?
+
+\noindent The Gene Ontology is quite well established in the field, so you will find GO annotations for almost all relevant UniProt entries or use dedicated tools like AmiGO or QuickGO to examine a protein of interest.
 
 ## Data
-The data that serve as the input for this project are functional annotation datasets in the GAF file format^[\scriptsize http://geneontology.org/docs/go-annotation-file-gaf-format-2.0/] which we generated with our GOMAP pipeline^[\scriptsize https://dill-picl.org/projects/gomap].
-They are (or will be shortly) available from \scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.
+Annotating genes with their functions can be done experimentally (e.g.
+by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
+Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner (CITE PAPER).
+We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
+These annotation sets are (or will be shortly) available from  
+\scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.
 
@@
 
-Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm.
+Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm (provided by PHYLIP).
+
+## Result
+
+The tree generated by neighbor joining is displayed in Figure \ref{fig:result}.
+4 of the 5 maize assemblies are grouped together in a clade, while the last one is an outlier to the remaining grasses and sits on a very long branch.
+In the remaining plants, cotton is an outlier to the legumes, which are all grouped in a single clade.
+
+\begin{figure*}[htbp]
+\includegraphics{resources/result.png}
+\caption{Resulting tree. maize\_v3 and maize\_v4 are two assemblies of the same maize cultivar, medicago1 and medicago2 are two different genotypes.}
+\label{fig:result}
+\end{figure*}
+
+## Discussion
+
+This tree shows surprising similarity to the taxonimic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.].
+The maize cultivars are all grouped together and there's even the distinction between the non-stiff stalk (W22 and Mo17), iodent (Ph207), and stiff stalk (maize_v4) classes.
+Only maize_v3, which is the same cultivar as maize_v4 (B73) is waaaay out there where it probably doesn't belong.
+We're currently investigating this but it seems that there is a lot wrong with the genome assembly itself, which would naturally influence the functional annotations derived from it.
+
+## Reproducing the Tree
+To reproduce this tree, follow these steps on a Linux machine that has PHYLIP available:
+
+```bash
+git clone https://github.com/Thyra/EEOB563.git paper_dennis # Clone the repository
+cd paper_dennis/final_project # change into the directory
+./build_distance_matrix > infile
+module load phylip # (if you're on HPC)
+neighbor
+```
+
+\noindent The resulting tree is in `outtree`.
+Manually root it (e.g. with Dendrogram) outside of the grasses (maize, wheat, barley) and you should end up with the same tree.
+This will use the pre-summarized `.tree.json` files in annotation_sets, if you want to completely reproduce it from scratch, delete them and only leave the `.gaf.gz` files.
+But be warned, this will take a while to calculate (probably over two hours).

### 2019-04-19 final_project/paper/Paper.Rmd
Final Draft stage

@@
 `r newthought("When we build phylogenetic trees,")` we often use the sequences of certain genes as the basis for our tree.
-They are the direct substrate of evolution so that makes sense, but when we think of evolution on a macro level, we don't really think of the sequences, we think of phenotype.
-Using phenotypes for evolutionary trees is a hard thing to do though because the choice of characteristics can be kind of arbitrary (http://wiki.c2.com/?BorgesClassificationOfAnimals Chinesische Tiere (Jorge Luis Borges 1966)), because it is very far away from the material evolution actually happens on, i.e.
-DNA.
-In this paper, I tried to go a middle ground: Instead of looking at the genetic sequences or the macromolecular phenotype, I'm looking at the presence of certain biological processes, molecular functions, or cellular components, in other words: along the path of evolution, any meaningful change in an organism goes along with the development, change, or loss of such a function; what good is a change in the DNA sequence if it does not lead to a modified function of that gene?
+They are the direct substrate of evolution so that makes sense, but practically often too abstract to be helpful.
+Let's say for example, that you have a patient infected with a novel pathogen.
+You can identify that pathogen and place it somewhere in the bacterial taxonomy, but unfortunately, none of the treatments that you usually use for closely related species show any effect against this one.
+How come?
+Just because two organisms are closely related does not *necessarily* mean that they are similar in their phenotype.
+So a tree that groups organisms based on their actual *in vivo* similarity would be much more helpful here.
+There are such trees built on phenotypic characteristics but unfortunately the choice of what characteristics to look at is not trivial and always situation-dependent.
+They can also be considered quite arbitrary\sidenote{
+Look at this (parodic) taxonomy of animals by Jorge Luis Borges:
+1. those that belong to the Emperor,
+2. embalmed ones,
+3. those that are trained,
+4. suckling pigs,
+5. mermaids,
+6. fabulous ones,
+7. stray dogs,
+8. those included in the present classification,
+9. those that tremble as if they were mad,
+10. innumerable ones,
+11. those drawn with a very fine camelhair brush,
+12. others,
+13. those that have just broken a flower vase,
+14. those that from a long way off look like flies.
+}.
+What's needed would be something that is exhaustive, clearly defined, and reproducible, but also more meaningful than just a DNA sequence.
+And they did it [@Zhu2015]!
+
+In this paper, I'm going to do something slightly similar. Not on bacteria, but on plants though.
+I know, disapointing, but I'll find a similarly fascinating reason why that's important soon.
+So, to summarize: I want to build a phylogenetic tree on plant species not based on their genetic sequence but on the *functions* that this plant is able to execute.
 
@@
 by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
-Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner (CITE PAPER).
+Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner [@Wimalanathan2018].
 We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
@@
 Applying this formula to all pairwise combinations of the genomes we're looking at yields a $S \times S$ distance matrix that can then serve as the input for a neighbor joining algorithm (provided by PHYLIP).
+I rooted the resulting tree manually outside of the grasses (maize, wheat, rice, barley).
 
@@
 
-The tree generated by neighbor joining is displayed in Figure \ref{fig:result}.
+The phylogram is displayed in Figure \ref{fig:result}.
 4 of the 5 maize assemblies are grouped together in a clade, while the last one is an outlier to the remaining grasses and sits on a very long branch.
@@
 \includegraphics{resources/result.png}
-\caption{Resulting tree. maize\_v3 and maize\_v4 are two assemblies of the same maize cultivar, medicago1 and medicago2 are two different genotypes.}
+\caption{Resulting phylogram. maize\_v3 and maize\_v4 are two assemblies of the same maize cultivar, medicago1 and medicago2 are two different genotypes.}
 \label{fig:result}
@@
 
-This tree shows surprising similarity to the taxonimic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.].
+This tree shows surprising similarity to the taxonimic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.]:
 The maize cultivars are all grouped together and there's even the distinction between the non-stiff stalk (W22 and Mo17), iodent (Ph207), and stiff stalk (maize_v4) classes.
 Only maize_v3, which is the same cultivar as maize_v4 (B73) is waaaay out there where it probably doesn't belong.
-We're currently investigating this but it seems that there is a lot wrong with the genome assembly itself, which would naturally influence the functional annotations derived from it.
+We're currently investigating this but it seems that a lot went wrong with the genome assembly itself, which would naturally influence the functional annotations derived from it.
+The two medicago genotypes, while correctly placed together in one clade, are an outlier to the other legumes.
+In a taxonomic tree, their place would be switched with peanut.
+Steven Cannon, who is an evolutionary biologist working on legumes, assumes that the reason for that is that medicago has a much higher rate of evolution and gene birth and death than other plants [@Young2011].
+As you will have noticed, this is all just a work in progress at the moment.
+My PI and I will meet with Steven on Monday and discuss (among other things) the following questions -- feel free to comment in your review if you have any input!
+
+- How similar is this tree to the taxonomic tree?
+- What are the reasons for the differences, what are the reasons for the similarities?
+- What biases in the construction process need we be aware of?
+- What does the tree depict/how do we interpret it?
+- In what way could it be scientifically valuable?
+- Are there any other taxa that would be good to have in the tree?
+- Is there any further analysis that would be good to do?
+(e.g.
+branch support, looking at each GO aspect separately, a more sophisticated measure of GAF similarity, trying a character state based method instead of distance...)
+- Is rooting the tree there reasonable?
 
@@
 module load phylip # (if you're on HPC)
-neighbor
+neighbor # (use standard options)
 ```

### 2019-04-22 final_project/visualize_makefile.sh
Restructuring, added Makefile visualization

@@ -1,0 +1,3 @@
+make -Bnd | ./make2graph > makefile.dot
+dot -omakevile.svg -Tsvg makefile.dot
+rm makefile.dot

### 2019-04-23 final_project/src/create_obo_jsons.py
Restructuring and added tool to find unique terms

@@ -1,5 +1,5 @@
-# This script reads an OBO file using the obo_parser and outputs a JSON file that is then used by the crystal scripts.
-# The JSON file looks like this:
+# This script reads an OBO file using the obo_parser and outputs JSON files that are then used by the crystal scripts.
+# The slimGO.JSON file looks like this:
 # {
 #   "parents": {
 #     "GO:122335": ["GO:4433", "GO:5432"],
@@ -9,6 +9,13 @@
 #   "main_id" : {"GO:1443":"GO:532", "GO:2533":"GO:5234"}
 #   "obsolete": ["GO:1432", "GO:4322", ...]
 # }
+#
+# The go_names.json file looks like this:
+# {
+#   "GO:43334":"Photosynthesis",
+#   "GO:93743":"Something else",
+#   ...
+# }
 
 import obo_parser
 import json
@@ -16,7 +23,8 @@ import json
 parents = {}
 main_id = {}
 obsolete = []
-with open("go.obo") as obofile:
+go_names = {}
+with open("resources/go.obo") as obofile:
   parser = obo_parser.Parser(obofile)
   for stanza in parser:
     if not stanza.name == "Term":
@@ -35,7 +43,11 @@ with open("go.obo") as obofile:
 
     if 'is_obsolete' in stanza.tags and stanza.tags['is_obsolete'][0] == "true":
       obsolete.append(go_id)
+    else:
+      go_names[go_id] = stanza.tags['name'][0]
 
 with open('slimGO.json', 'w') as f:
   f.write(json.dumps({"parents":parents,"main_id":main_id,"obsolete":obsolete},indent=2))
 
+with open('go_names.json', 'w') as f:
+  f.write(json.dumps(go_names,indent=2))

### 2019-04-24 final_project/annotation_sets/go_counts/count.sh
Added cowpea and script to count GO terms

@@ -1,0 +1,4 @@
+while read term; do
+  echo $term
+  zgrep -c "$term" *.gaf.gz > go_counts/$term.count
+done < go_counts/terms

### 2019-04-25 final_project/paper/Paper.Rmd
worked on paper

@@
 
-The Gene Ontology is such an ontology that describes genes by the properties of their product.
+The Gene Ontology (GO) is such an ontology that describes genes by the properties of their product.
 \marginnote{"The mission of the GO Consortium is to develop an up-to-date, comprehensive, computational model of biological systems, from the molecular level to larger pathways, cellular and organism-level systems."
@@
 \includegraphics{resources/result.png}
-\caption{Resulting phylogram. maize\_v3 and maize\_v4 are two assemblies of the same maize cultivar, medicago1 and medicago2 are two different genotypes.}
+\caption{Phylogram built on the Jaccard distance matrix with Neighbor Joining. Manually rooted between monocots and dicots and text in green added.}
 \label{fig:result}
@@
 
+The tree is surprisingly similar to the taxonomic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.], but why did it work so well?
+Given the number of taxa, it is unlikely that it is due to pure chance or noise but it is not clear whether this tree actually demonstrates a biologically meaningful difference in the functional annotations of these species or wehther it is simply a reproduction of some bias or artifact that stems from some element in the process.
+
+To further examine this question, I wanted to see where the phylogenetic signal actually comes from.
+So I asked the following question: Which terms are common to all nitrogen fixing plants but do not occur in any other plant in the tree?^[in mathy words this is the intersection of terms in the nitrogen fixing species minus the union of terms in all other species]
+If these terms are actually meaningful to the process of nitrogen fixation (or some other characteristic that clearly differentiates this group of plants from the others) that would indicate that the phylogenetic signal actually comes from biologically plausible differences in the functional annotations.
+So, here is the answer:
+
+- GO:0080184 - response to phenylpropanoid
+- GO:0033800 - isoflavone 7-O-methyltransferase activity
+- GO:0042577 - lipid phosphatase activity
+- GO:0031174 - lifelong otolith mineralization
+- GO:0045299 - otolith mineralization
+- GO:0006742 - NADP catabolic process
+- GO:0019364 - pyridine nucleotide catabolic process
+- GO:0019677 - NAD catabolic process
+- GO:0070823 - HDA1 complex
+
+Three of these terms do seem plausible: The Nod Factor molecule that is secreted by the plant as a signal to Rhizobia bacteria is a phenylpropanoid, isoflavone 7-O-methyltransferase activity is involved in Nod Factor synthesis, and the nodule has a complex lipid membrane system – and Nod Factor has a lipid component (which could explain lipid phosphatase activity).
+The last 4 terms are possibly plausible, but less certain, but the two otolith terms seem very out of place, since they describe small oval calcareous bodies in the inner ear of vertebrates, not something expected to be found in legumes.
+
 ## Reproducing the Tree

### 2019-04-25 final_project/src/visualize_makefile.sh
Restructured things, added parsimony methods

@@ -1,3 +1,0 @@
-make -Bnd | ./make2graph > makefile.dot
-dot -omakevile.svg -Tsvg makefile.dot
-rm makefile.dot

### 2019-04-25 final_project/paper/Paper.Rmd
Added Parsimony method and result to paper

@@
 
-## The Gene Ontology
+# The Gene Ontology
 Historically, the function/role that a gene plays in an organism has always been described in natural language, however the researcher characterizing that gene deemed best.
@@
 
-## Data
+# Data
 Annotating genes with their functions can be done experimentally (e.g.
@@
 
-## Method
-Starting point of our method are the functional annotation sets, one for each genome, which annotate every gene in the genome with one or more GO terms.
+# Method
+I am using two different tree building approaches, one is a distanced based method and the other a parsimony one.
+
+## Distance Based
+Starting point of our analysis are the functional annotation sets, one for each genome, which annotate every gene in the genome with one or more GO terms.
 In more mathematical terms the genome annotation set is a list of tuples $(G, T)$ with $G \in$ Genes in that genome and $T \in$ Terms in the Gene Ontology.
@@
 
-## Result
+## Parsimony Based
 
-The phylogram is displayed in Figure \ref{fig:result}.
-4 of the 5 maize assemblies are grouped together in a clade, while the last one is an outlier to the remaining grasses and sits on a very long branch.
-In the remaining plants, cotton is an outlier to the legumes, which are all grouped in a single clade.
+Like in the previous method, we again start by enriching all taxa sets $(G,T)$ with their ancestor terms and discarding the gene association: $S = \bigcup_{i=1}^{x}(T_i \cup A_{i})$.
+
+Instead of using these sets for distance calculation, we combine them into a big binary matrix that displays which terms are present in which set:
+
+\begin{table}[]
+\begin{tabular}{lllllll}
+Taxon       & \scriptsize{GO:0016021} & \scriptsize{GO:0009699} & \scriptsize{GO:0050794} & \scriptsize{GO:0050789} & \scriptsize{GO:0060739} & ... \\
+\textit{G. max}      & 1          & 0          & 1          & 0          & 1          &     \\
+\textit{T. aestivum} & 1          & 1          & 0          & 1          & 1          &     \\
+...         &            &            &            &            &            &    
+\end{tabular}
+\end{table}
+
+This matrix was then used as the input for `pars` from the PHYLIP package to find the maximally parsimonious tree.
+
+# Results
+
+The phylogram created by the distance based method is displayed in Figure \ref{fig:result}, the maximum parsimony tree in figure \ref{fig:parsimony}.
 
@@
 
-## Discussion
+\begin{figure*}[htbp]
+\includegraphics{resources/parsimonyTree.png}
+\caption{Phylogram built by looking for the maximally parsimonious tree (total of 7780 changes). Manually rooted between monocots and dicots.}
+\label{fig:parsimony}
+\end{figure*}
+
+Both trees were combined into a tanglegram (see figure \ref{fig:tanglegram}).
+
+\begin{figure*}[htbp]
+\includegraphics{resources/tanglegram.png}
+\caption{Tanglegram of the distance based and maximum parsimony tree (both manually rooted).}
+\label{fig:tanglegram}
+\end{figure*}
+
+# Discussion
 

### 2019-04-28 final_project/paper/Paper.Rmd
New Discussion

@@
 \begin{figure*}[htbp]
-\includegraphics{resources/tanglegram.png}
+\includegraphics{resources/Tanglegram.png}
 \caption{Tanglegram of the distance based and maximum parsimony tree (both manually rooted).}
@@
 
-This tree shows surprising similarity to the taxonimic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.]:
-The maize cultivars are all grouped together and there's even the distinction between the non-stiff stalk (W22 and Mo17), iodent (Ph207), and stiff stalk (maize_v4) classes.
-Only maize_v3, which is the same cultivar as maize_v4 (B73) is waaaay out there where it probably doesn't belong.
-We're currently investigating this but it seems that a lot went wrong with the genome assembly itself, which would naturally influence the functional annotations derived from it.
-The two medicago genotypes, while correctly placed together in one clade, are an outlier to the other legumes.
-In a taxonomic tree, their place would be switched with peanut.
-Steven Cannon, who is an evolutionary biologist working on legumes, assumes that the reason for that is that medicago has a much higher rate of evolution and gene birth and death than other plants [@Young2011].
-As you will have noticed, this is all just a work in progress at the moment.
-My PI and I will meet with Steven on Monday and discuss (among other things) the following questions -- feel free to comment in your review if you have any input!
+## Fulfilling expectancies?
+In a perfect world, the evolution of organisms could be just as easily retraced by the evolution of their *functions* as by their sequence; after all the selective pressure is on the mechinisms and activities of an organism and not on its DNA sequence.
+\marginnote{Although you may get lower resolution, because silent mutations or such that don't alter the function of a protein would be missed.}
 
-- How similar is this tree to the taxonomic tree?
-- What are the reasons for the differences, what are the reasons for the similarities?
-- What biases in the construction process need we be aware of?
-- What does the tree depict/how do we interpret it?
-- In what way could it be scientifically valuable?
-- Are there any other taxa that would be good to have in the tree?
-- Is there any further analysis that would be good to do?
-(e.g.
-branch support, looking at each GO aspect separately, a more sophisticated measure of GAF similarity, trying a character state based method instead of distance...)
-- Is rooting the tree there reasonable?
+The expected taxonimic tree is displayed in figure \ref{fig:expectedTree} and indeed the tree produced by the distance based method is quite similar (much more similar than I had expected).
+There are only two notable differences: *Sorghum* should be at the base of maize and not grouped with *Brachypodium* and *Medicago* and *Arachis* should switch their places.
+We are currently investigating the reason for this strange phenomenon and the best explanatioi we've come up with so far is that *Brachy* and *Medicago* are not actually representative of their respective group; they have been chosen as model organisms because they are easy to sequence but that's mainly because their genome is much smaller and less complex than that of actual crop plants.
+*Medicago* is also well known for having an unusally high rate of genomic evolution, gene births and gene deaths [@Young2011].
 
-The tree is surprisingly similar to the taxonomic tree one might expect if it had been built on sequences^[I should probably display that expected tree here.], but why did it work so well?
-Given the number of taxa, it is unlikely that it is due to pure chance or noise but it is not clear whether this tree actually demonstrates a biologically meaningful difference in the functional annotations of these species or wehther it is simply a reproduction of some bias or artifact that stems from some element in the process.
+\begin{marginfigure}
+\includegraphics{resources/expectedTree.png}
+\caption{Expected cladogram according to sequence based phylogenetics. \citep{Kellogg2001,Azani2017,Puttick2018}}
+\label{fig:expectedTree}
+\end{marginfigure}
 
-To further examine this question, I wanted to see where the phylogenetic signal actually comes from.
-So I asked the following question: Which terms are common to all nitrogen fixing plants but do not occur in any other plant in the tree?^[in mathy words this is the intersection of terms in the nitrogen fixing species minus the union of terms in all other species]
+It is difficult to come to a clear evolutionary tree within the same species (I am talking about maize^[Many of the maize lines have more than one ancestor and there is an immense amount of horizontal gene transfer with maize, even to closely related species.]) but it is encouraging to see that in the generated tree the 3 different classes of maize (stiff stalk, non-stiff stalk, and iodent) are differentiated from each other in a reasonable way.
+
+`r newthought("What about the parsimony tree?")` Surprisingly, the tree built on the parsimony method is not identical with the distance based one, and it is even less similar to the expected tree.
+Our first thought was that since the parsimony method does not search the complete tree space, it might simply not have found the distance based tree even though it is the most parsimonious one.
+Unfortunately that does not seem to be the case: The tree it found has a score of 7780, the distance based tree 7821.
+When manually switching *Medicago* and *Arachis* to get closer to the taxonomic tree, the score improved to 7812, but putting *Sorghum* at the place it was expected increased the score again to 7866.
+So it seems the answer is not that simple.
+
+## Tracing Back the Signal
+Another big question is why does the tree actually look the way it does?
+Is it truly a biologically meaningful display of differences and similarities in function or is it just an artifactual tree, mainly caused by a bias in the method.
+Given the number of taxa it seems unlikely that the similarity to the expected tree is just random noise, so there must be some systematic reason behind it.
+To make a first step answering this question, I investigated what actually causes species to cluster the way they do. I asked the following question: 
+Which terms are common to all nitrogen fixing plants but do not occur in any other plant in the tree?^[in mathy words this is the intersection of terms in the nitrogen fixing species minus the union of terms in all other species]
 If these terms are actually meaningful to the process of nitrogen fixation (or some other characteristic that clearly differentiates this group of plants from the others) that would indicate that the phylogenetic signal actually comes from biologically plausible differences in the functional annotations.
@@
 The last 4 terms are possibly plausible, but less certain, but the two otolith terms seem very out of place, since they describe small oval calcareous bodies in the inner ear of vertebrates, not something expected to be found in legumes.
+There might still be some biological significance to it but we're not sure yet what that might be.
+So for now it seems that some of the signal clearly comes from biological differences while some other part might be due to the method.
+
+## What's next?
+
+Since this is part of a publication I am working on, a lot more thought will be put into, for example, the following questions:
+
+- How similar is this tree to the taxonomic tree?
+- What are the reasons for the differences, what are the reasons for the similarities?
+- What biases in the construction process need we be aware of?
+- What does the tree depict/how do we interpret it?
+- In what way could it be scientifically valuable?
+- Are there any other taxa that would be good to have in the tree?
+- Is there any further analysis that would be good to do? (e.g. branch support, looking at each GO aspect separately, a more sophisticated measure of GAF similarity...)
+- Is rooting the tree there reasonable?
 

### 2019-04-29 final_project/paper/Paper.Rmd
Updated reproducing section and added reviews

@@
 author: "Dennis Psaroudakis"
-date: "April 18th 2019"
+date: "April 28th 2019"
 keep_md: yes
@@
 
-## Reproducing the Tree
+# Reproducing the Tree
 To reproduce this tree, follow these steps on a Linux machine that has PHYLIP available:
@@
 cd paper_dennis/final_project # change into the directory
-./build_distance_matrix > infile
+
+# Distance based method
+bin/build_distance_matrix annotation_sets/*tree.json > distance_matrix.phy
 module load phylip # (if you're on HPC)
 neighbor # (use standard options)
+
+# Parsimony based Method
+bin/binary_table annotation_sets/*.tree.json > binary.csv
+bin/binary2phylip binary.csv > binary.phy
+module load phylip
+pars # (again, standard options)
+
+# If you're adding more GAFs
+bin/gaf2json <new_gaf>
+# Then repeat distance or parsimony as above
+
+# To replace GAF filenames with species names in newick tree
+bin/rename_taxa <tree.newick> resources/taxa_name_mapping.csv
 ```
 
-\noindent The resulting tree is in `outtree`.
-Manually root it (e.g. with Dendrogram) outside of the grasses (maize, wheat, barley) and you should end up with the same tree.
-This will use the pre-summarized `.tree.json` files in annotation_sets, if you want to completely reproduce it from scratch, delete them and only leave the `.gaf.gz` files.
-But be warned, this will take a while to calculate (probably over two hours).
+\noindent If any of the binaries fails without a message that says otherwise, you probably need to install some more libraries.
+If that doesn't help you may have to recompile them for your processor type (they're written in [Crystal](https://crystal-lang.org/), source code is in `src/`)

### 2019-04-29 final_project/paper/Paper.Rmd
Rewrote intro

@@
 
-`r newthought("When we build phylogenetic trees,")` we often use the sequences of certain genes as the basis for our tree.
-They are the direct substrate of evolution so that makes sense, but practically often too abstract to be helpful.
-Let's say for example, that you have a patient infected with a novel pathogen.
-You can identify that pathogen and place it somewhere in the bacterial taxonomy, but unfortunately, none of the treatments that you usually use for closely related species show any effect against this one.
+`r newthought("When we build phylogenetic trees,")` we often use the DNA or protein sequences of certain genes^[with appropriate degree of conservedness depending on the evolutionary distance of the taxa, little probability of horizontal gene transfer, reliable for sequencing and alignment etc.] as the basis for our analysis.
+They are the direct substance of evolution so that makes sense, especially when we need high resolution (i.e. a signal in very closely related organisms).
+In some cases, they don't give the answer we're looking for though:
+Let's say for example, that you're a doctor and one of your patients is infected with a novel pathogen.
+You can sequence that pathogen and place it somewhere in the bacterial taxonomy, and naturally you will start treating the patient based on what you know about closely related species.
+Unfortunately, none of the antibiotics you're using show any effect.
 How come?
 Just because two organisms are closely related does not *necessarily* mean that they are similar in their phenotype.
-So a tree that groups organisms based on their actual *in vivo* similarity would be much more helpful here.
-There are such trees built on phenotypic characteristics but unfortunately the choice of what characteristics to look at is not trivial and always situation-dependent.
-They can also be considered quite arbitrary\sidenote{
+Especially if you've only used genes that are suitable for regular phylogenetics but don't actually have anything to do with the organism's metabolism, membrane structure etc, you have no guarantee that close evolutionary relationship means similar response to environmental factors.
+In this case, a tree that can group your pathogen with organisms based on their actual *in vivo* similarity would be much more helpful here.
+There are trees built on phenotypic characteristics (nose color, ear length...) but the choice of what characteristics to look at is not trivial and has an incredible influence on the result.
+You will never be able to cover all relevant attributes and your choice can therefore be considered somewhat arbitrary\sidenote{
 Look at this (parodic) taxonomy of animals by Jorge Luis Borges:
@@
 13. those that have just broken a flower vase,
-14. those that from a long way off look like flies.
+14. those that from a long way off look like flies.\medskip
 }.
-What's needed would be something that is exhaustive, clearly defined, and reproducible, but also more meaningful than just a DNA sequence.
-And they did it [@Zhu2015]!
+A method would be needed that is exhaustive, clearly defined, and reproducible, but also more effective at finding the answer you're looking for than using the DNA sequence.
+And, unsurprisingly, there is [@Zhu2015] (it's a cool paper, check it out)!
 
-In this paper, I'm going to do something slightly similar. Not on bacteria, but on plants though.
-I know, disapointing, but I'll find a similarly fascinating reason why that's important soon.
-So, to summarize: I want to build a phylogenetic tree on plant species not based on their genetic sequence but on the *functions* that this plant is able to execute.
+In this paper, I'm going to do something less exciting. Instead of working with life-threatening bacteria and saving patients, I will focus on plants.
+The idea is somewhat similar though^[and I had the idea myself first and only found the paper afterwards]: I want to build a phylogenetic tree of plant species, not based on their genetic sequence but on the processes, structures, and reactions that are present in it.
 
 # The Gene Ontology
-Historically, the function/role that a gene plays in an organism has always been described in natural language, however the researcher characterizing that gene deemed best.
+Historically, the function/role that a gene plays in an organism has been described in natural language, however the researcher characterizing that gene deemed best.
 While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind the words.
@@
 \includegraphics{resources/expectedTree.png}
-\caption{Expected cladogram according to sequence based phylogenetics. \citep{Kellogg2001,Azani2017,Puttick2018}}
+\caption{Expected cladogram. \citep{Kellogg2001,Azani2017,Puttick2018}}
 \label{fig:expectedTree}
+\medskip
 \end{marginfigure}

### 2019-04-29 final_project/paper/Paper.Rmd
Rewrite some more stuff

@@
 While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind the words.
-Additionally, different people will describe the same thing with different words, which has the potential for misunderstanding.
+Additionally, different people will describe the same thing with different words, so there is always a potential for misunderstanding.
 
@@
 
-The Gene Ontology (GO) is such an ontology that describes genes by the properties of their product.
+The Gene Ontology (GO) is such an ontology.
+It describes genes by the properties of their product.
 \marginnote{"The mission of the GO Consortium is to develop an up-to-date, comprehensive, computational model of biological systems, from the molecular level to larger pathways, cellular and organism-level systems."
@@
 
-- How similar in function is protein A to protein B? (One answer would be: How many steps in the GO graph do I need from term A to B? The fewer steps, the more similar the function)
+- How similar in function^[*Function* refers to any aspect of the GO, not just Molecular Function.] is protein A to protein B?
+(One answer would be: How many steps in the GO graph do I need from term A to B?
+The fewer steps, the more similar the function)
 - If protein A is involved in Biological Process XYZ, what other proteins are involved in that same process?
 
-\noindent The Gene Ontology is quite well established in the field, so you will find GO annotations for almost all relevant UniProt entries or use dedicated tools like AmiGO or QuickGO to examine a protein of interest.
+\noindent The Gene Ontology is quite well established in the field, so you will find GO annotations all across UniProt entries or be able to use dedicated tools like [AmiGO](http://amigo.geneontology.org/amigo) or [QuickGO](https://www.ebi.ac.uk/QuickGO/) to examine a protein of interest.
 
@@
 Annotating genes with their functions can be done experimentally (e.g.
-by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
-Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner [@Wimalanathan2018].
+by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try to predict the function of a given gene.
+Our lab has developed such a pipeline called [Gene Ontology Meta Annotator for Plants (GOMAP)](https://dill-picl.org/projects/gomap/) which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner [@Wimalanathan2018].
 We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
@@
 ## Fulfilling expectancies?
-In a perfect world, the evolution of organisms could be just as easily retraced by the evolution of their *functions* as by their sequence; after all the selective pressure is on the mechinisms and activities of an organism and not on its DNA sequence.
+In a perfect world, the evolution of organisms could be just as easily retraced by the evolution of their functions as by their sequence; after all the selective pressure is on the mechinisms and activities of an organism and not on its DNA sequence and what differentiates two species from each other are the differences in their *in vivo* pheontype.
 \marginnote{Although you may get lower resolution, because silent mutations or such that don't alter the function of a protein would be missed.}
@@
 There are only two notable differences: *Sorghum* should be at the base of maize and not grouped with *Brachypodium* and *Medicago* and *Arachis* should switch their places.
-We are currently investigating the reason for this strange phenomenon and the best explanatioi we've come up with so far is that *Brachy* and *Medicago* are not actually representative of their respective group; they have been chosen as model organisms because they are easy to sequence but that's mainly because their genome is much smaller and less complex than that of actual crop plants.
+We are currently investigating the reason for this phenomenon and the best explanation we've come up with so far is that *Brachy* and *Medicago* are not actually good representations of their respective group; they have been chosen as model organisms because they are easy to sequence but that's mainly because their genome is much smaller and less complex than that of actual crop plants.
 *Medicago* is also well known for having an unusally high rate of genomic evolution, gene births and gene deaths [@Young2011].
@@
 
-It is difficult to come to a clear evolutionary tree within the same species (I am talking about maize^[Many of the maize lines have more than one ancestor and there is an immense amount of horizontal gene transfer with maize, even to closely related species.]) but it is encouraging to see that in the generated tree the 3 different classes of maize (stiff stalk, non-stiff stalk, and iodent) are differentiated from each other in a reasonable way.
+It is difficult to come up with a clear evolutionary tree within the same species (I am talking about maize^[Many of the maize lines have more than one ancestor and there is an immense amount of horizontal gene transfer with maize, even to closely related species.]) but it is encouraging to see that in the generated tree the 3 different classes of maize (stiff stalk, non-stiff stalk, and iodent) are differentiated from each other in a reasonable way.
 
@@
 When manually switching *Medicago* and *Arachis* to get closer to the taxonomic tree, the score improved to 7812, but putting *Sorghum* at the place it was expected increased the score again to 7866.
-So it seems the answer is not that simple.
+So it seems the answer is not that simple and we will have to reflect more on our methods to understand where the difference comes from and which is the more meaningful result.
 
 ## Tracing Back the Signal
-Another big question is why does the tree actually look the way it does?
-Is it truly a biologically meaningful display of differences and similarities in function or is it just an artifactual tree, mainly caused by a bias in the method.
+Another big question is why the tree actually looks the way it does.
+Is it truly a biologically meaningful display of differences and similarities in function or is it just an artifactual tree, caused by a bias in the method.
 Given the number of taxa it seems unlikely that the similarity to the expected tree is just random noise, so there must be some systematic reason behind it.
@@
 There might still be some biological significance to it but we're not sure yet what that might be.
-So for now it seems that some of the signal clearly comes from biological differences while some other part might be due to the method.
+So for now it seems that some of the signal clearly comes from biological differences while some other part might be due to the method.^[I looked at other, similar constellations as well and the answer was the same (e.g.
+what differentiates non-stiff stalk maize lines from the others).]
 
@@
 - Are there any other taxa that would be good to have in the tree?
+- Should we get rid of *Brachypodium* and *Medicago*?
 - Is there any further analysis that would be good to do? (e.g. branch support, looking at each GO aspect separately, a more sophisticated measure of GAF similarity...)
